Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North

### 2023-04-20 analysis/paper/paper.qmd
ready to submit

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University"
author:
  - Valentin Lucet:
      # correspondence: "yes"
      email: valentin.lucet@gmail.com
      orcid: "0000-0003-0268-818X"
      institute: wlu
  - Samantha McFarlane:
      institute: eccc
  - Jennifer Baltzer:
      institute: wlu
  - Cheryl A. Johnson:   
      institute: eccc
  - Mathieu Leblond:
      institute: eccc
  - Eric Neilson:
      institute: nrcan
  - Philip Weibe:
      institute: nrcan
  - Frances Stewart:
      institute: wlu
      orcid: "0000-0001-9344-8346"
  - Josie Hughes:
      institute: eccc
institute:
  - wlu:
      name: Wilfrid Laurier University
  - eccc:
      name: Environment & Climate Change Canada
  - nrcan:
      name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template_modified.docx" # Insert path for the DOCX file
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
abstract: |
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage. We continue to use this workflow for ongoing image identification while investigating requirements to develop a machine learning image classifier.

@@

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Documents/WILDLab/rofcamtrap/')
backup_options <- options()
options(scipen = 1, digits = 12)
library(magrittr)
```

@@

Out of the __`r processed_fmt`__ detections, __`r dels_nb_fmt`__ were false positives (or __`r dels_nb_pct` %__), where Megadetector identified something, but there was nothing. In addition, __`r non_mans_nb_fmt`__ were true positives (or __`r non_mans_nb_pct` %__), to which __`r mans_nb_fmt`__ false negatives were added (manually added bounding boxes, __`r mans_nb_pct` %__). Out of the __`r non_mans_nb_fmt`__ true positives, __`r orig_tb_fmt[3]`__ had to have their bounding box adjusted. The below summarises information for the tagging progress.

@@

We obtained repeat detections of at least 31 mammal and bird species across 193 sites and 5-7 months of detections The most common species detected (detected more than 10 days in the season) are listed in Table 1 (the full table is available in the appendix at the end of this report). Note the large amount detections in the “bird” and “duck” species categories which could be further distinguished into specific species.

@@

sp_tab %>%
  dplyr::left_join(sci_names, by="species") %>% 
  dplyr::relocate(Scientific_name, .after=species) %>% 
  dplyr::filter(days_detected >= 10) %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Relative abundances for species with more than 10 detection days",
    subtitle = "Computed in number of days detected in the season"
  ) %>%
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) %>% 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
    Scientific_name = "Scientific name"
  ) %>% 
  gt::tab_style(style = list(
      gt::cell_text(style="italic")),
    locations = gt::cells_body(columns = Scientific_name)) %>% 
  gt::opt_stylize(style = 6, color = "gray")

@@

![](../figures/Caribou_bplot.png)
![](../figures/Sandhill crane_bplot.png)
![](../figures/Moose_bplot.png)

![](../figures/Wolf_bplot.png)

@@

Over the coming months we will continue to process the retrieved data and anticipate adding to this data set over the summer, as data from ## cameras has yet to be retrieved. The next steps for this project are to complete the tagging process and identify all the detections tagged as unknowns, bird, duck and mustelid species. We will also further explore the spatial community composition patterns at each camera site. Given a high enough data density for certain species future possibilities for this project include modelling species occupancy (@MacKenzie2002) and spatially explicit density estimation using spatially explicit capture-recapture models (@Chandler2013).

@@

sp_tab %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Relative species abundances",
    subtitle = "Computed in number of days detected in the season"
  ) %>%
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) %>% 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
  ) %>% 
  gt::opt_stylize(style = 6, color = "gray")
```

@@

lu_dat %>% 
  dplyr::select(category_code, label) %>% 
  dplyr::filter(category_code != "ND") %>% 
  dplyr::arrange(category_code) %>% 
  gt::gt() %>%
  gt::tab_header(
    title = "Land cover classes abbreviations corresponding to the Ontatio Far North Land cover dataset.",
    subtitle = "Adapted from the ontario Far North land cover dataset"
  ) %>%
  gt::cols_label(
    label ="Land cover class",
    category_code = "Abbreviations",
  ) %>% 
  gt::opt_stylize(style = 1, color = "gray")
```

@@

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

>>>>

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University"
author:
  - Valentin Lucet:
      # correspondence: "yes"
      email: valentin.lucet@gmail.com
      orcid: "0000-0003-0268-818X"
      institute: wlu
  - Samantha McFarlane:
      institute: eccc
  - Jennifer Baltzer:
      institute: wlu
  - Cheryl A. Johnson:   
      institute: eccc
  - Mathieu Leblond:
      institute: eccc
  - Eric Neilson:
      institute: nrcan
  - Philip Weibe:
      institute: nrcan
  - Frances Stewart:
      institute: wlu
      orcid: "0000-0001-9344-8346"
  - Josie Hughes:
      institute: eccc
institute:
  - wlu:
      name: Wilfrid Laurier University
  - eccc:
      name: Environment & Climate Change Canada
  - nrcan:
      name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format:
  # docx:
  #   reference-doc: "../templates/template_modified.docx" # Insert path for the DOCX file
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    extra_dependencies: ["float"]
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
abstract: |
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage. We continue to use this workflow for ongoing image identification while investigating requirements to develop a machine learning image classifier.

@@

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Documents/WILDLab/rofcamtrap/')
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
backup_options <- options()
options(scipen = 1, digits = 2)
library(magrittr)
```

@@

Out of the __`r processed_fmt`__ detections, __`r dels_nb_fmt`__ were false positives (or __`r dels_nb_pct` %__), where Megadetector identified something, but there was nothing. In addition, __`r non_mans_nb_fmt`__ were true positives (or __`r non_mans_nb_pct` %__), to which __`r mans_nb_fmt`__ false negatives were added (manually added bounding boxes, __`r mans_nb_pct` %__). Out of the __`r non_mans_nb_fmt`__ true positives, __`r orig_tb_fmt[3]`__ had to have their bounding box adjusted. The table below summarises information for the tagging progress.

@@

We obtained repeat detections of at least 31 mammal and bird species across 193 sites and 5-7 months of detections The most common species detected (detected more than 10 days in the season) are listed in Table 2 below (the full table is available in the appendix at the end of this report). Note the large amount detections in the “bird” and “duck” species categories which could be further distinguished into specific species.

@@

sp_tab %>%
  dplyr::left_join(sci_names, by="species") %>% 
  dplyr::relocate(Scientific_name, .after=species) %>% 
  dplyr::filter(days_detected >= 10) %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Table 2: Relative abundances for species with more than 10 detection days",
    subtitle = "Computed in number of days detected in the season"
  ) %>%
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) %>% 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
    Scientific_name = "Scientific name"
  ) %>% 
  gt::tab_style(style = list(
      gt::cell_text(style="italic")),
    locations = gt::cells_body(columns = Scientific_name)) %>% 
  gt::opt_stylize(style = 6, color = "gray")

@@

![](../figures/Caribou_bplot.png)
\vspace{1pt}

![](../figures/Sandhill crane_bplot.png)
\vspace{1pt}

![](../figures/Moose_bplot.png)
\vspace{1pt}

![](../figures/Wolf_bplot.png)
\vspace{1pt}

@@

Over the coming months we will continue to process the retrieved data and anticipate adding to this data set over the summer, as data from 18 cameras has yet to be retrieved. The next steps for this project are to complete the tagging process and identify all the detections tagged as unknowns, bird, duck and mustelid species. We will also further explore the spatial community composition patterns at each camera site. Given a high enough data density for certain species future possibilities for this project include modelling species occupancy (@MacKenzie2002) and spatially explicit density estimation using spatially explicit capture-recapture models (@Chandler2013).

@@

sp_tab %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Table 3: Relative species abundances",
    subtitle = "Computed in number of days detected in the season"
  ) %>%
  gt::fmt_number(
    decimals = 0,
    columns = days_detected,
    suffixing = TRUE
  ) %>% 
  gt::cols_label(
    days_detected ="Days",
    species = "Species",
  ) %>% 
  gt::opt_stylize(style = 6, color = "gray")
```

@@

lu_dat %>% 
  dplyr::select(category_code, label) %>% 
  dplyr::filter(category_code != "ND") %>% 
  dplyr::arrange(category_code) %>% 
  gt::gt() %>%
  gt::tab_header(
    title = "Table 4: Land cover classes abbreviations corresponding to the Ontatio Far North Land cover dataset.",
    subtitle = "Adapted from the ontario Far North land cover dataset"
  ) %>%
  gt::cols_label(
    label ="Land cover class",
    category_code = "Abbreviations",
  ) %>% 
  gt::opt_stylize(style = 1, color = "gray")
```

@@

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & 
    git2r::in_repository(path = ".")) {
  git2r::repository(here::here())  
}
```
